---
title: "BikeShare"
author: "Zelibe_Samuel"
date: "2024-06-20"
output: html_document
---

## Installing required packages

```{r}
install.packages("tidyverse")
install.packages("janitor")
install.packages("lubridate")
install.packages("ggplot2")
```

## Loading the packages

```{r}
library(tidyverse)
library(janitor)
library(lubridate)
library(ggplot2)
library(tidyr)
library(readr)
```

## Import data files and rename to identify easily

```{r}
jul23 <- read_csv("case1/202306-divvy-tripdata.csv")
aug23 <- read_csv("case1/202307-divvy-tripdata.csv")
sep23 <- read_csv("case1/202308-divvy-tripdata.csv")
oct23 <- read_csv("case1/202309-divvy-tripdata.csv")
nov23 <- read_csv("case1/202310-divvy-tripdata.csv")
dec23 <- read_csv("case1/202311-divvy-tripdata.csv")
jan24 <- read_csv("case1/202312-divvy-tripdata.csv")
feb24 <- read_csv("case1/202401-divvy-tripdata.csv")
mar24 <- read_csv("case1/202402-divvy-tripdata.csv")
apr24 <- read_csv("case1/202403-divvy-tripdata.csv")
may24 <- read_csv("case1/202404-divvy-tripdata.csv")
jun24 <- read_csv("case1/202405-divvy-tripdata.csv")
```

## Check the structure of the imported data

```{r}
str(jul23)
str(aug23)
str(sep23)
str(oct23)
str(nov23)
str(dec23)
str(jan24)
str(feb24)
str(mar24)
str(apr24)
str(may24)
str(jun24)
```

```{r}
col(jul23)
col(aug23)
col(sep23)
col(oct23)
col(nov23)
col(dec23)
col(jan24)
col(feb24)
col(mar24)
col(apr24)
col(may24)
col(jun24)
```

## Concatenating all the data into a single data frame

```{r}
bike_trips <- rbind(jul23,aug23,sep23,oct23,nov23,dec23,jan24,feb24,mar24,apr24,may24,jun24)
```

#How many rows are in the data frame

```{r}
nrow(bike_trips)
```

#Data dimension

```{r}
dim(bike_trips)
```

#The first six rows

```{r}
head(bike_trips)
```

## Adding columns that list date,month,day and year of each ride

```{r}
bike_trips$date <- as.Date(bike_trips$started_at)  #default yyyy-mm-dd
bike_trips$month <- format(as.Date(bike_trips$date), "%m")
bike_trips$day <- format(as.Date(bike_trips$date), "%d")
bike_trips$year <- format(as.Date(bike_trips$date), "%y")
bike_trips$day_of_week <- format(as.Date(bike_trips$date), "%A")
```

#Check for new column names

```{r}
colnames(bike_trips)
View(bike_trips)
```

#A new column will be added to compare the ride length for each ride

```{r}
bike_trips$ride_length <- difftime(bike_trips$ended_at, bike_trips$started_at)
```

## Remove data with ride length less than 0 second and is greater than 1440 mins should be removed as its either negative or more than one day

```{r}
new_bike_trips <- bike_trips[!(bike_trips$ride_length <= 0 | bike_trips$ride_length > 1440),]
dim(new_bike_trips)
```

## Removing NA"s

```{r}
new_bike_trips <- drop_na(new_bike_trips)
dim(new_bike_trips)
summary(new_bike_trips)
```

##Remove duplicates

```{r}
new_bike_trips_1 <- new_bike_trips[!duplicated(new_bike_trips$ride_id),]
dim(new_bike_trips_1)
```

## Assign the correct order to each day of the week

```{r}
new_bike_trips_1$day_of_week <- ordered(new_bike_trips_1$day_of_week, levels = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday', 'Sunday'))

new_bike_trips_1 %>% group_by(member_casual, day_of_week) %>% 
  summarise(number_of_ride = n(), .groups = 'drop') %>% 
  arrange(day_of_week)
```

##Lets assign the correct order to each month of the year

```{r}
new_bike_trips_1$month <- ordered(new_bike_trips_1$month, levels =c('06', '07', '08', '09', '10', '11', '12', '01', '02', '03', '04', '05'))

new_bike_trips_1 %>% group_by(member_casual, month) %>% 
  summarise(number_of_ride = n(), .groups = 'drop') %>% 
  arrange(month)
```

## Different ride_length depending on rider type

```{r}
aggregate(new_bike_trips_1$ride_length ~ new_bike_trips_1$member_casual + new_bike_trips_1$day_of_week, FUN = mean)

new_bike_trips_1 %>%  group_by(member_casual, month) %>% 
  summarise(average_ride_length = mean(ride_length), .groups = 'drop') %>% arrange(month)
```

## MAX

```{r}
aggregate(new_bike_trips_1$ride_length ~ new_bike_trips_1$member_casual, FUN = max)
```

## MEDIAN

```{r}
aggregate(new_bike_trips_1$ride_length ~ new_bike_trips_1$member_casual, FUN = median)
```

## MIN

```{r}
aggregate(new_bike_trips_1$ride_length ~ new_bike_trips_1$member_casual, FUN = min)
```

**This indicates that members make shorter trips in contrast to casual riders. While casual riders spend more time on the road, members tend to ride the bikes more frequently**

## Checking trip duration to know how each rider use the bike and sharing visualization

```{r}
new_bike_trips_1 %>% drop_na(member_casual) %>% group_by(member_casual, day_of_week) %>% 
  summarise(average_trip_duration = mean(ride_length), .groups = 'keep') %>% 
  ggplot(aes(x = day_of_week, y = average_trip_duration, fill = member_casual)) + geom_col(width = 0.5, position = position_dodge(width = 0.5)) + labs(title = "Average trip duration by rider type Vs. Day of the week")
```

## Lets count by month

```{r}
new_bike_trips_1 %>%  drop_na(member_casual) %>%  group_by(member_casual, month) %>% 
  summarise(number_of_rides = n(), .groups = 'keep') %>% 
  arrange(member_casual, month) %>%
  
  ggplot(aes(x = month, y = number_of_rides, fill = member_casual)) + labs(title = "Total trips by rider type vs. month") + theme(axis.text.x = element_text(angle = 30)) + 
  geom_col(width = 0.5, position = position_dodge(width = 0.5)) +
  scale_y_continuous(labels = function(x) format(x, scientific =FALSE))
```

##Average trip duration

```{r}
new_bike_trips_1 %>% drop_na(member_casual) %>% group_by(member_casual, month) %>% 
  summarise(average_trip_duration = mean(ride_length), .groups = 'keep') %>% 
  ggplot(aes(x = month, y = average_trip_duration, fill = member_casual)) + geom_col(width = 0.5, position = position_dodge(width = 0.5)) + labs(title = "Average trip duration by rider type Vs. month")
```

## lets check for the rideable type

```{r}
ride_type <- new_bike_trips_1 %>% ggplot(aes(x = member_casual, fill = rideable_type)) + 
  geom_bar() +
  theme_minimal() + 
  theme(axis.text.x  = element_text(angle = 45)) +
  labs(title = "Types of Bikes Used",
       subtitle = "Period: 2022/09 - 2023/08",
       x = "Type of Rider",
       y = "Count")
ride_type + scale_fill_brewer(palette = "Set1", name = " ",
                              labels = c("Classic bike",
                                         "Docked bike",
                                         "Electric bike")) +
  scale_x_discrete(labels = c("Casual", "Member"))
```

**From the chart above it shows that both members and casual riders use classic bikes and electric bikes. while only casual riders use docked bikes**

## Now im going to compare how many trips each group has taken

```{r}
new_bike_trips_1%>%  
  drop_na(member_casual) %>%
  group_by(member_casual) %>%
  count() %>%
  ungroup() %>%
  mutate(perc = `n` / sum(`n`)) %>%
  arrange(perc) %>%
  mutate(labels = scales::percent(perc)) %>%
  ggplot(aes(x="", 
             y=perc, 
             fill=member_casual))  +
  geom_bar(stat="identity", width=1, color = "white") +
  geom_text(aes(label = labels), color = "white",
            position = position_stack(vjust = 0.5)) +
  coord_polar("y", start=0) +
  theme_void()
```

**This chart shows that members take more trips than casual riders significantly**
